---
title: "From a clinical question to an auditable variable"
output:
  rmarkdown::html_vignette:
    css: extractionengine.css
vignette: >
  %\VignetteIndexEntry{From a clinical question to an auditable variable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
options(tibble.width = Inf)
```

`extractionengine` executes a study variable that **you** define. It does not
choose the clinical definition, interpret missing evidence as clinical absence,
or decide which source column is the result.

This vignette follows one small, fully synthetic example. By the end, you will
be able to:

- separate a reusable concept from a variable-specific activation;
- read `combine by -> filter by qualified -> group by -> evaluate value` as a relational
  data flow;
- predict which rows reach an output expression; and
- find the published value, selection and processing state, evidence, and audit
  trail.

The analyte codes, values, and thresholds below are illustrative. They are not a
recommended clinical definition.

```{r libraries, message = FALSE, warning = FALSE}
library(extractionengine)
library(dplyr)
library(tibble)
```

## 1. Begin with the row universe

The cohort is the set of units the study asks the engine to evaluate. Here it
contains four stays belonging to two patients. The biology table is a prepared
source: in a real study, `redsan` owns that preparation step.

`ELTID` is the required generic element coordinate. Biology may also carry its
native `BIOL_ID`; the engine preserves it when present but does not require it.
Both identifiers are local to the biology source: an `ELTID` must never be joined
to an `ELTID` from a different source.

```{r synthetic-data}
cohort <- tribble(
  ~PATID, ~EVTID,
  "P1",   "E1",
  "P1",   "E2",
  "P2",   "E3",
  "P2",   "E4"
)

biology <- tribble(
  ~PATID, ~EVTID, ~ELTID, ~DATEXAM,    ~TYPEANA, ~NUMRES, ~STRRES,
  "P1",   "E1",   "L1",   "2026-01-01", "K.K",       5.8, NA_character_,
  "P1",   "E1",   "L2",   "2026-01-02", "K.K",       5.6, NA_character_,
  "P1",   "E1",   "L3",   "2026-01-02", "NA.NA",   130,   NA_character_,
  "P1",   "E2",   "L4",   "2026-02-01", "K.K",       4.0, NA_character_,
  "P1",   "E2",   "L5",   "2026-02-02", "K.K",       4.2, NA_character_,
  "P1",   "E2",   "L6",   "2026-02-02", "NA.NA",   140,   NA_character_,
  "P2",   "E3",   "L7",   "2026-03-01", "K.K",       5.5, NA_character_,
  "P2",   "E3",   "L8",   "2026-03-01", "NA.NA",   141,   NA_character_,
  "P2",   "E4",   "L9",   "2026-04-01", "K.K",       5.9, NA_character_,
  "P2",   "E4",   "L10",  "2026-04-01", "NA.NA",   129,   NA_character_
) |>
  mutate(DATEXAM = as.Date(DATEXAM))

cohort
biology |>
  select(PATID, EVTID, DATEXAM, TYPEANA, NUMRES)
```

Before reading further, inspect the rows and try to answer two questions:

1. Which stays have both potassium above 5.2 and sodium below 135?
2. For each patient, how would the mean potassium change if it used only those
   stays rather than every stay of a patient who qualifies at least once?

## 2. A concept only locates possible evidence

Keep reusable concepts atomic. Potassium and sodium each declare where their own
rows are found; a variable can activate both without creating a composite
"ionic balance" concept. `analyte()` matches `TYPEANA`; it does not choose
`NUMRES`, `STRRES`, or `DATEXAM` as the result.

```{r concept}
potassium <- concept_spec(
  name = "potassium",
  channels = list(
    lab = lab_channel(
      selector = analyte("K.K")
    )
  )
)

sodium <- concept_spec(
  name = "sodium",
  channels = list(
    lab = lab_channel(
      selector = analyte("NA.NA")
    )
  )
)
```

The names connect the authoring layers explicitly:

- `lab_channel()` uses the logical source name `"biology"` by default;
  `sources = list(biology = biology)` supplies that source at runtime;
- both concepts may use the local channel name `lab`; and
- inside a variable, the name on the left of `channels = list(...)` is an
  activation alias, while
  `use_channel(channel = "lab", concept = potassium)` names both the reusable
  concept and the channel it activates.

Combine expressions, outputs, evidence, and audit tables refer to activation
aliases. An alias may reuse its origin's name, as in the first example, or give
the origin a variable-specific role such as `potassium_high`. The concept object
belongs to the activation rather than the whole `variable_spec()`, which is what
allows one variable to compose independently reusable concepts.

The authoring layers answer different questions:

| Layer | Question |
|---|---|
| `concept_spec()` | Where can relevant source rows be found? |
| `use_channel()` | How does this variable use those rows? |
| `combine_channels()` | Which relational units satisfy the signal rule? |
| `from_channel()` / `bin_output()` | What is published, at which key, and how is its value calculated? |

## 3. Activate one channel and publish one value

Start with a simple question: *what is the mean potassium for each stay?*

```{r simple-variable}
mean_potassium_by_stay <- variable_spec(
  name = "mean_potassium_by_stay",

  channels = list(
    potassium = use_channel(
      channel = "lab",
      concept = potassium
    )
  ),

  output = from_channel(
    channel = "potassium",
    group_by = "EVTID",
    value = mean(NUMRES, na.rm = TRUE)
  )
)

potassium_run <- run_variable(
  mean_potassium_by_stay,
  cohort = cohort,
  sources = list(biology = biology)
)

potassium_run$values
```

Read the output declaration from top to bottom:

- take the activated channel called `potassium`;
- group its remaining source rows by `EVTID`; and
- evaluate `mean(NUMRES, na.rm = TRUE)` once for each group.

The runner calls each final output unit a **task**. A task is not specific to an
LLM: with `group_by = "EVTID"`, the engine creates one task per stay; with
`group_by = "PATID"`, it creates one task per patient.

For a deterministic channel, `value` is one data-masked expression: bare names
refer to aligned prepared-source columns. It sees missing values too, so handling
them with `na.rm`, `is.na()`, or another explicit rule belongs to the author. The
expression must return exactly one atomic value or one list cell per task. With
no candidate rows it is not evaluated and the engine publishes logical `NA`.

## 4. Qualify stays without publishing a laboratory value

Now define an observed alert as potassium above 5.2 **and** sodium below 135 in
the same stay. `filter_rows` has dplyr `filter()` semantics and is evaluated
separately inside each task: its data-masked expression returns one logical value
per candidate row, with missing results treated as false. Each activation keeps
the complete surviving source rows; the combine then tests whether both signals
coexist at `EVTID`.

The expressions use the same data-mask vocabulary as dplyr. Bare `NUMRES` is the
usual form; `.data$NUMRES` makes the source-column lookup explicit and
`.env$threshold` reads an object from the authoring environment.

```{r binary-variable}
ionic_alert_by_stay <- variable_spec(
  name = "ionic_alert_by_stay",

  channels = list(
    potassium_high = use_channel(
      channel = "lab",
      concept = potassium,
      filter_rows = NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "lab",
      concept = sodium,
      filter_rows = NUMRES < 135
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = bin_output(group_by = "EVTID")
)
```

`combine_channels(..., by = "EVTID")` determines **where qualification is
decided**. It does not calculate a mean or choose the final output key.
`bin_output(group_by = "EVTID")` publishes the observed membership directly:
one `1` or `0` for each cohort stay.

## 5. Qualification and output rows are separate decisions

Suppose the scientific question changes:

> Among patients with at least one alert stay, what is their mean potassium?

That sentence is incomplete. It can mean either:

- mean potassium from the alert stays only; or
- mean potassium from all stays of a patient who has at least one alert stay.

The combine is identical in both cases. A third activation, `potassium_rows`, is
payload-only: it supplies values to the output without participating in the
boolean expression.

The cohort still declares which `PATID + EVTID` stays are eligible. The output's
`group_by = "PATID"` projects those eligible units to one final task per patient;
the source `EVTID` remains available to the finer combine relation.

First, retain only potassium rows belonging to alert stays:

```{r qualified-stays-variable}
mean_k_in_qualified_stays <- variable_spec(
  name = "mean_k_in_qualified_stays",

  channels = list(
    potassium_high = use_channel(
      channel = "lab",
      concept = potassium,
      filter_rows = NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "lab",
      concept = sodium,
      filter_rows = NUMRES < 135
    ),
    potassium_rows = use_channel(
      channel = "lab",
      concept = potassium
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = from_channel(
    channel = "potassium_rows",
    group_by = "PATID",
    value = mean(NUMRES, na.rm = TRUE),
    filter_by_qualified = "EVTID"
  )
)
```

The second specification is deliberately repeated so that its complete contract
remains visible. Only its canonical name and `filter_by_qualified` change:

```{r qualified-patients-variable}
mean_k_in_qualified_patients <- variable_spec(
  name = "mean_k_in_qualified_patients",

  channels = list(
    potassium_high = use_channel(
      channel = "lab",
      concept = potassium,
      filter_rows = NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "lab",
      concept = sodium,
      filter_rows = NUMRES < 135
    ),
    potassium_rows = use_channel(
      channel = "lab",
      concept = potassium
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = from_channel(
    channel = "potassium_rows",
    group_by = "PATID",
    value = mean(NUMRES, na.rm = TRUE),
    filter_by_qualified = "PATID"
  )
)
```

Run the three related variables together. `run_protocol()` returns a lossless
named list; each result name comes from `variable_spec(name = ...)`.

```{r run-protocol}
results <- run_protocol(
  list(
    ionic_alert_by_stay,
    mean_k_in_qualified_stays,
    mean_k_in_qualified_patients
  ),
  cohort = cohort,
  sources = list(biology = biology)
)

results$ionic_alert_by_stay$values

bind_rows(
  results$mean_k_in_qualified_stays$values,
  results$mean_k_in_qualified_patients$values
) |>
  select(PATID, variable, value)
```

The execution order explains the difference:

1. `combine by EVTID`: identify alert stays;
2. `filter by qualified EVTID` or `PATID`: choose which native potassium rows
   may enter the output;
3. `group by PATID`: form one final group per patient;
4. `evaluate value`: calculate one cell from each group's remaining columns.

There is no hidden mean per stay. The `value` expression is evaluated once at
the final `group_by` key.

For `P1`, filtering by `EVTID` keeps only the two potassium values from alert
stay `E1`, so the result is `(5.8 + 5.6) / 2 = 5.7`. Filtering by `PATID` keeps
all four potassium values from `E1` and `E2`, so the result becomes `4.9`.

## 6. The same distinction in dplyr

The engine declaration is compact, but its relational meaning should be
translatable to familiar data manipulation.

```{r dplyr-equivalent}
potassium_high <- biology |>
  filter(TYPEANA == "K.K", NUMRES > 5.2) |>
  distinct(PATID, EVTID)

sodium_low <- biology |>
  filter(TYPEANA == "NA.NA", NUMRES < 135) |>
  distinct(PATID, EVTID)

qualified_evtids <- potassium_high |>
  inner_join(sodium_low, by = c("PATID", "EVTID"))

potassium_rows <- biology |>
  filter(TYPEANA == "K.K")

# filter_by_qualified = "EVTID"
potassium_rows |>
  semi_join(qualified_evtids, by = c("PATID", "EVTID")) |>
  group_by(PATID) |>
  summarise(value = mean(NUMRES, na.rm = TRUE), .groups = "drop")

# filter_by_qualified = "PATID"
qualified_patients <- qualified_evtids |>
  distinct(PATID)

potassium_rows |>
  semi_join(qualified_patients, by = "PATID") |>
  group_by(PATID) |>
  summarise(value = mean(NUMRES, na.rm = TRUE), .groups = "drop")
```

`filter_by_qualified` therefore chooses the key used by the conceptual
`semi_join()` **before** terminal grouping. It does not choose the level at which
the output expression operates; `output$group_by` does that.

## 7. Read the result as an audit package

Every run has four top-level components:

```{r result-components}
names(results$mean_k_in_qualified_stays)
```

### What was published?

`values` is the study variable at the declared output key.

```{r audit-values}
results$mean_k_in_qualified_stays$values
```

### Did execution finish, and was a signal observed?

`channel_status` separates row selection from optional model processing.

```{r audit-status}
results$ionic_alert_by_stay$channel_status |>
  select(PATID, EVTID, channel, selection_status, processing_status)
```

`selection_status` is `matched`, `no_match`, or `unavailable`. The first two say
that selection could be evaluated and did or did not retain candidates; the
third says that no reliable selection was possible. These deterministic channels
need no model-processing stage, so `processing_status = "not_required"`. An LLM
channel instead uses `completed`, `not_called`, `invalid`, or `failed`. None of
these labels is a clinical interpretation.

### Which source rows remain inspectable?

`evidence` contains the source occurrences retained by the executed rule. Here
we inspect only the payload rows that reached the patient result after
qualification.

```{r audit-evidence}
results$mean_k_in_qualified_stays$evidence |>
  filter(channel == "potassium_rows") |>
  select(PATID, source_EVTID, evidence_kind, TYPEANA, NUMRES, evidence_ref)
```

The final output key is `PATID`, while `source_EVTID` records the stay to which a
source row natively belonged. The two identifiers answer different questions in
a cross-grain result. At event-grain output, target `EVTID` and `source_EVTID`
remain distinct roles even when their values are equal. `evidence_kind`
distinguishes a prepared `source_row`, a `lucene_hit`, and an `llm_citation`.
`evidence_ref` is an opaque coordinate local to this run and source snapshot,
not a permanent warehouse identifier; private executor coordinates are kept out
of this table.

### What recipe ran, and why did a stay qualify?

The execution manifest is the resolved configuration snapshot, not a
chronological log.

```{r audit-manifest}
manifest_lines <- capture.output(
  print(results$mean_k_in_qualified_stays$audit$execution_manifest)
)

# Omit only the run timestamp so that the rendered vignette is reproducible.
cat(manifest_lines[!startsWith(manifest_lines, "Executed at:")], sep = "\n")
```

`combine_keys` exposes the boolean relation evaluated at `combine$by` for keys
observed in at least one combine activation. It is not a complete cohort roster:
an all-false stay may be absent here even though `bin_output()` publishes its
final `0` in `values`.

```{r audit-combine-keys}
results$mean_k_in_qualified_stays$audit$combine_keys
```

`counts` helps locate where rows were removed. Its controlled stages are
`pre_selector`, `window`, `selector`, `filtered_selector`, `model_input`, and
`output_input`: they follow rows from relational eligibility through selection,
activation filters, optional model input, and terminal value input. They count
structured source rows, searchable documents, or snippets, not clinical
outcomes.

```{r audit-counts}
results$mean_k_in_qualified_stays$audit$counts |>
  filter(PATID == "P1", channel == "potassium_rows")
```

Low-level executor objects live under `audit$internal`. They are useful for
debugging, but they are not the normal onboarding surface.

## 8. A checklist for authoring a new variable

Before running a new definition, you should be able to answer these questions in
plain language:

1. Which cohort keys form the row universe?
2. Which source rows does each concept channel locate?
3. Which row or group filters does each activation apply?
4. At which key must signals coexist?
5. Which qualified rows feed the published value?
6. Which key defines one final output row?
7. Which expression calculates one cell from the remaining aligned columns?

As small exercises, change one declaration at a time and predict the result
before rerunning:

- publish the ionic alert once per `PATID` rather than per `EVTID`;
- change the combine from `&` to `|`; or
- replace `mean(NUMRES, na.rm = TRUE)` with `max(NUMRES, na.rm = TRUE)` and
  explain which rows the expression sees.

Text retrieval and structured LLM extraction use the same
concept--activation--output separation, but add retrieval scope, prompts, and an
`ellmer::TypeObject`. They are intentionally kept out of this deterministic
onboarding path. Continue with
[`vignette("structured-text-with-llm", package = "extractionengine")`](structured-text-with-llm.html)
for the LLM workflow. See `?use_channel` for windows, grouped filters, and text
search.
