---
title: "From a clinical question to an auditable variable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a clinical question to an auditable variable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
options(tibble.width = Inf)
```

`extractionengine` executes a study variable that **you** define. It does not
choose the clinical definition, interpret missing evidence as clinical absence,
or decide which source column is the result.

This vignette follows one small, fully synthetic example. By the end, you will
be able to:

- separate a reusable concept from a variable-specific activation;
- read `combine by -> filter by qualified -> group by -> reduce` as a relational
  data flow;
- predict which rows reach a reducer; and
- find the published value, execution status, evidence, and audit trail.

The analyte codes, values, and thresholds below are illustrative. They are not a
recommended clinical definition.

```{r libraries, message = FALSE, warning = FALSE}
library(extractionengine)
library(dplyr)
library(tibble)
```

## 1. Begin with the row universe

The cohort is the set of units the study asks the engine to evaluate. Here it
contains four stays belonging to two patients. The biology table is a prepared
source: in a real study, `redsan` owns that preparation step.

```{r synthetic-data}
cohort <- tribble(
  ~PATID, ~EVTID,
  "P1",   "E1",
  "P1",   "E2",
  "P2",   "E3",
  "P2",   "E4"
)

biology <- tribble(
  ~PATID, ~EVTID, ~ELTID, ~biol_ID, ~DATEXAM,    ~TYPEANA, ~NUMRES, ~STRRES,
  "P1",   "E1",   "L1",   "B1",     "2026-01-01", "K.K",       5.8, NA_character_,
  "P1",   "E1",   "L2",   "B2",     "2026-01-02", "K.K",       5.6, NA_character_,
  "P1",   "E1",   "L3",   "B3",     "2026-01-02", "NA.NA",   130,   NA_character_,
  "P1",   "E2",   "L4",   "B4",     "2026-02-01", "K.K",       4.0, NA_character_,
  "P1",   "E2",   "L5",   "B5",     "2026-02-02", "K.K",       4.2, NA_character_,
  "P1",   "E2",   "L6",   "B6",     "2026-02-02", "NA.NA",   140,   NA_character_,
  "P2",   "E3",   "L7",   "B7",     "2026-03-01", "K.K",       5.5, NA_character_,
  "P2",   "E3",   "L8",   "B8",     "2026-03-01", "NA.NA",   141,   NA_character_,
  "P2",   "E4",   "L9",   "B9",     "2026-04-01", "K.K",       5.9, NA_character_,
  "P2",   "E4",   "L10",  "B10",    "2026-04-01", "NA.NA",   129,   NA_character_
) |>
  mutate(DATEXAM = as.Date(DATEXAM))

cohort
biology |>
  select(PATID, EVTID, DATEXAM, TYPEANA, NUMRES)
```

Before reading further, inspect the rows and try to answer two questions:

1. Which stays have both potassium above 5.2 and sodium below 135?
2. For each patient, how would the mean potassium change if it used only those
   stays rather than every stay of a patient who qualifies at least once?

## 2. A concept only locates possible evidence

The concept declares where potassium and sodium rows are found. `analyte()`
matches `TYPEANA`; it does not choose `NUMRES`, `STRRES`, or `DATEXAM` as the
result.

```{r concept}
ionic_balance <- concept_spec(
  name = "ionic_balance",
  channels = list(
    potassium = lab_channel(
      selector = analyte("K.K")
    ),
    sodium = lab_channel(
      selector = analyte("NA.NA")
    )
  )
)
```

The names connect the authoring layers explicitly:

- `lab_channel()` uses the logical source name `"biology"` by default;
  `sources = list(biology = biology)` supplies that source at runtime;
- `potassium` and `sodium`, on the left of the concept's `channels` list, are
  concept-channel names; and
- inside a variable, the name on the left of `channels = list(...)` is an
  activation alias, while `use_channel(channel = "potassium")` points back to
  the concept channel it activates.

Combine expressions, outputs, evidence, and audit tables refer to activation
aliases. An alias may reuse its origin's name, as in the first example, or give
the same origin a variable-specific role such as `potassium_high`.

The authoring layers answer different questions:

| Layer | Question |
|---|---|
| `concept_spec()` | Where can relevant source rows be found? |
| `use_channel()` | How does this variable use those rows? |
| `combine_channels()` | Which relational units satisfy the signal rule? |
| `from_channel()` / `bin_output()` | What is published, at which key, and how is it reduced? |

## 3. Activate one channel and publish one value

Start with a simple question: *what is the mean potassium for each stay?*

```{r simple-variable}
mean_potassium_by_stay <- variable_spec(
  name = "mean_potassium_by_stay",
  concept = ionic_balance,

  channels = list(
    potassium = use_channel(
      channel = "potassium"
    )
  ),

  output = from_channel(
    channel = "potassium",
    column = "NUMRES",
    group_by = "EVTID",
    reduce = mean
  )
)

potassium_run <- run_variable(
  mean_potassium_by_stay,
  cohort = cohort,
  sources = list(biology = biology)
)

potassium_run$values
```

Read the output declaration from top to bottom:

- take the activated channel called `potassium`;
- read its real prepared-source column `NUMRES`;
- group the remaining rows by `EVTID`; and
- call `mean()` once for each group.

The runner calls each final output unit a **task**. A task is not specific to an
LLM: with `group_by = "EVTID"`, the engine creates one task per stay; with
`group_by = "PATID"`, it creates one task per patient.

The reducer is optional. With `reduce = NULL`, zero non-missing values produce a
typed `NA`, one value is published directly, and more than one is a cardinality
error. Supplying `reduce = mean` explicitly defines how multiple values become
one scalar.

## 4. Qualify stays without publishing a laboratory value

Now define an observed alert as potassium above 5.2 **and** sodium below 135 in
the same stay. `filter_rows` has dplyr `filter()` semantics and is evaluated
separately inside each task: its named source columns arrive as vectors, and it
returns one logical value per candidate row. Each activation keeps the complete
surviving source rows; the combine then tests whether both signals coexist at
`EVTID`.

```{r binary-variable}
ionic_alert_by_stay <- variable_spec(
  name = "ionic_alert_by_stay",
  concept = ionic_balance,

  channels = list(
    potassium_high = use_channel(
      channel = "potassium",
      filter_rows = function(NUMRES) NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "sodium",
      filter_rows = function(NUMRES) NUMRES < 135
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = bin_output(group_by = "EVTID")
)
```

`combine_channels(..., by = "EVTID")` determines **where qualification is
decided**. It does not calculate a mean or choose the final output key.
`bin_output(group_by = "EVTID")` publishes the observed membership directly:
one `1` or `0` for each cohort stay.

## 5. Qualification and reducer input are separate decisions

Suppose the scientific question changes:

> Among patients with at least one alert stay, what is their mean potassium?

That sentence is incomplete. It can mean either:

- mean potassium from the alert stays only; or
- mean potassium from all stays of a patient who has at least one alert stay.

The combine is identical in both cases. A third activation, `potassium_rows`, is
payload-only: it supplies values to the output without participating in the
boolean expression.

The cohort still declares which `PATID + EVTID` stays are eligible. The output's
`group_by = "PATID"` projects those eligible units to one final task per patient;
the source `EVTID` remains available to the finer combine relation.

First, retain only potassium rows belonging to alert stays:

```{r qualified-stays-variable}
mean_k_in_qualified_stays <- variable_spec(
  name = "mean_k_in_qualified_stays",
  concept = ionic_balance,

  channels = list(
    potassium_high = use_channel(
      channel = "potassium",
      filter_rows = function(NUMRES) NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "sodium",
      filter_rows = function(NUMRES) NUMRES < 135
    ),
    potassium_rows = use_channel(
      channel = "potassium"
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = from_channel(
    channel = "potassium_rows",
    column = "NUMRES",
    filter_by_qualified = "EVTID",
    group_by = "PATID",
    reduce = mean
  )
)
```

The second specification is deliberately repeated so that its complete contract
remains visible. Only its canonical name and `filter_by_qualified` change:

```{r qualified-patients-variable}
mean_k_in_qualified_patients <- variable_spec(
  name = "mean_k_in_qualified_patients",
  concept = ionic_balance,

  channels = list(
    potassium_high = use_channel(
      channel = "potassium",
      filter_rows = function(NUMRES) NUMRES > 5.2
    ),
    sodium_low = use_channel(
      channel = "sodium",
      filter_rows = function(NUMRES) NUMRES < 135
    ),
    potassium_rows = use_channel(
      channel = "potassium"
    )
  ),

  combine = combine_channels(
    "potassium_high & sodium_low",
    by = "EVTID"
  ),

  output = from_channel(
    channel = "potassium_rows",
    column = "NUMRES",
    filter_by_qualified = "PATID",
    group_by = "PATID",
    reduce = mean
  )
)
```

Run the three related variables together. `run_protocol()` returns a lossless
named list; each result name comes from `variable_spec(name = ...)`.

```{r run-protocol}
results <- run_protocol(
  list(
    ionic_alert_by_stay,
    mean_k_in_qualified_stays,
    mean_k_in_qualified_patients
  ),
  cohort = cohort,
  sources = list(biology = biology)
)

results$ionic_alert_by_stay$values

bind_rows(
  results$mean_k_in_qualified_stays$values,
  results$mean_k_in_qualified_patients$values
) |>
  select(PATID, variable, value)
```

The execution order explains the difference:

1. `combine by EVTID`: identify alert stays;
2. `filter by qualified EVTID` or `PATID`: choose which native potassium rows
   may enter the output;
3. `group by PATID`: form one final group per patient;
4. `reduce with mean`: calculate one scalar from each group's remaining values.

There is no hidden mean per stay. The reducer is called once at the final
`group_by` key.

For `P1`, filtering by `EVTID` keeps only the two potassium values from alert
stay `E1`, so the result is `(5.8 + 5.6) / 2 = 5.7`. Filtering by `PATID` keeps
all four potassium values from `E1` and `E2`, so the result becomes `4.9`.

## 6. The same distinction in dplyr

The engine declaration is compact, but its relational meaning should be
translatable to familiar data manipulation.

```{r dplyr-equivalent}
potassium_high <- biology |>
  filter(TYPEANA == "K.K", NUMRES > 5.2) |>
  distinct(PATID, EVTID)

sodium_low <- biology |>
  filter(TYPEANA == "NA.NA", NUMRES < 135) |>
  distinct(PATID, EVTID)

qualified_evtids <- potassium_high |>
  inner_join(sodium_low, by = c("PATID", "EVTID"))

potassium_rows <- biology |>
  filter(TYPEANA == "K.K")

# filter_by_qualified = "EVTID"
potassium_rows |>
  semi_join(qualified_evtids, by = c("PATID", "EVTID")) |>
  group_by(PATID) |>
  summarise(value = mean(NUMRES), .groups = "drop")

# filter_by_qualified = "PATID"
qualified_patients <- qualified_evtids |>
  distinct(PATID)

potassium_rows |>
  semi_join(qualified_patients, by = "PATID") |>
  group_by(PATID) |>
  summarise(value = mean(NUMRES), .groups = "drop")
```

`filter_by_qualified` therefore chooses the key used by the conceptual
`semi_join()` **before** terminal grouping. It does not choose the level at which
the reducer operates; `output$group_by` does that.

## 7. Read the result as an audit package

Every run has four top-level components:

```{r result-components}
names(results$mean_k_in_qualified_stays)
```

### What was published?

`values` is the study variable at the declared output key.

```{r audit-values}
results$mean_k_in_qualified_stays$values
```

### Did execution finish, and was a signal observed?

`channel_status` separates processing from signal membership.

```{r audit-status}
results$ionic_alert_by_stay$channel_status |>
  select(PATID, EVTID, channel, status, hit, contribution)
```

`status = "complete"` means ascertainment finished; it does not mean that the
channel hit. `hit` records the observed signal. A contribution labelled
`negative` means "ascertained without an observed hit", not a clinical negative
finding.

### Which source rows remain inspectable?

`evidence` contains public, resolvable source evidence. Here we inspect only the
payload rows that reached the patient result after qualification.

```{r audit-evidence}
results$mean_k_in_qualified_stays$evidence |>
  filter(channel == "potassium_rows") |>
  select(PATID, source_EVTID, TYPEANA, NUMRES, evidence_ref)
```

The final output key is `PATID`, while `source_EVTID` records the stay to which a
source row natively belonged. The two identifiers answer different questions in
a cross-grain result. `evidence_ref` is the canonical public reference assigned
to that evidence row; private executor coordinates are kept out of this table.

### What recipe ran, and why did a stay qualify?

The execution manifest is the resolved configuration snapshot, not a
chronological log.

```{r audit-manifest}
manifest_lines <- capture.output(
  print(results$mean_k_in_qualified_stays$audit$execution_manifest)
)

# Omit only the run timestamp so that the rendered vignette is reproducible.
cat(manifest_lines[!startsWith(manifest_lines, "Executed at:")], sep = "\n")
```

`combine_keys` exposes the boolean relation evaluated at `combine$by` for keys
observed in at least one combine activation. It is not a complete cohort roster:
an all-false stay may be absent here even though `bin_output()` publishes its
final `0` in `values`.

```{r audit-combine-keys}
results$mean_k_in_qualified_stays$audit$combine_keys
```

`counts` helps locate where rows were removed. Its stages count rows or values;
they are not clinical outcomes.

```{r audit-counts}
results$mean_k_in_qualified_stays$audit$counts |>
  filter(PATID == "P1", channel == "potassium_rows")
```

Low-level executor objects live under `audit$internal`. They are useful for
debugging, but they are not the normal onboarding surface.

## 8. A checklist for authoring a new variable

Before running a new definition, you should be able to answer these questions in
plain language:

1. Which cohort keys form the row universe?
2. Which source rows does each concept channel locate?
3. Which row or group filters does each activation apply?
4. At which key must signals coexist?
5. Which qualified rows feed the published value?
6. Which key defines one final output row?
7. What must the reducer return when it receives the remaining values?

As small exercises, change one declaration at a time and predict the result
before rerunning:

- publish the ionic alert once per `PATID` rather than per `EVTID`;
- change the combine from `&` to `|`; or
- replace `mean` with `max` and explain which rows it receives.

Text retrieval and structured LLM extraction use the same
concept--activation--output separation, but add retrieval scope, prompts, and an
`ellmer::TypeObject`. They are intentionally kept out of this deterministic
onboarding path. See `?use_channel` for windows, grouped filters, and text search,
and the README's
[Structured text extraction](https://github.com/FrancescoMonti-source/extraction-engine#structured-text-extraction)
section for the LLM contract.
