---
title: "Structured text extraction with an LLM"
output:
  rmarkdown::html_vignette:
    css: extractionengine.css
vignette: >
  %\VignetteIndexEntry{Structured text extraction with an LLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
options(tibble.width = Inf)
```

This vignette starts where
[`vignette("getting-started", package = "extractionengine")`](getting-started.html)
ends. It assumes that you already know the roles of `concept_spec()`,
`use_channel()`, and `from_channel()`. Here the new questions are:

- which documents may supply evidence for one output task;
- which excerpts are sent to the model;
- how to describe a structured response with `ellmer`;
- when a response is grounded enough to publish; and
- how to distinguish retrieval, model processing, evidence, and audit state.

The example is fully synthetic. The model call is deliberately not run while
the package vignette is built; every step before that call is executable and
checked offline.

```{r libraries, message = FALSE, warning = FALSE}
library(extractionengine)
library(dplyr)
library(tibble)
```

## 1. Start with one explicit extraction question

Suppose the study asks:

> For each stay, what medication-allergy status is explicitly documented, and
> which substance is named?

The wording matters. Silence must not become absence of allergy, and the model
must justify its record using only excerpts supplied by the engine.

The cohort defines three stay-level output tasks. The document source contains
one document per stay, but only two documents mention allergy.

```{r document-source, message = FALSE, warning = FALSE}
cohort <- tribble(
  ~PATID, ~EVTID,
  "P1",   "E1",
  "P2",   "E2",
  "P3",   "E3"
)

document_rows <- tribble(
  ~ELTID, ~RECTXT, ~PATID, ~EVTID, ~RECDATE, ~RECTYPE,
  "D1",
  "Allergie medicamenteuse documentee a la penicilline, avec urticaire.",
  "P1", "E1", "2026-01-04", "CR",
  "D2",
  "Le patient nie toute allergie medicamenteuse connue.",
  "P2", "E2", "2026-02-08", "CR",
  "D3",
  "Sortie a domicile avec traitement habituel inchange.",
  "P3", "E3", "2026-03-12", "CR"
) |>
  mutate(RECDATE = as.Date(RECDATE))

documents <- corpustools::create_tcorpus(
  as.data.frame(document_rows),
  text_columns = "RECTXT",
  doc_column = "ELTID",
  split_sentences = TRUE,
  remember_spaces = FALSE,
  verbose = FALSE
)
```

A metadata-rich `corpustools::tCorpus` is the canonical public document source.
`ELTID` identifies a native document inside that source; `PATID`, `EVTID`,
`RECDATE`, and `RECTYPE` remain available for relational scope and evidence.

## 2. Make retrieval inspectable before adding an LLM

The concept only locates text that may be relevant. A Lucene match is a
candidate, not the final allergy classification.

```{r text-concept}
medication_allergy <- concept_spec(
  name = "medication_allergy",
  channels = list(
    allergy_text = text_channel(
      selector = lucene_query("allerg*")
    )
  )
)
```

Before paying the cost of a model call, exercise the retrieval boundary with a
deterministic `method = "lucene"` variable.

```{r retrieval-check, message = FALSE, warning = FALSE}
allergy_mentions <- variable_spec(
  name = "allergy_mentions",
  concept = medication_allergy,
  channels = list(
    allergy_text = use_channel(
      channel = "allergy_text",
      search_within = "EVTID",
      method = "lucene"
    )
  ),
  output = bin_output(group_by = "EVTID")
)

retrieval_run <- run_variable(
  allergy_mentions,
  cohort = cohort,
  sources = list(documents = documents)
)

retrieval_run$values

retrieval_run$evidence |>
  select(
    PATID, EVTID, source_EVTID, evidence_kind,
    snippet_id, snippet_text
  )
```

This answers only: *did the query retain at least one excerpt in this stay?*
It deliberately does not answer whether an allergy is affirmed or denied. The
second document demonstrates why Lucene-hit presence cannot itself become the
clinical value.

`search_within = "EVTID"` means that each task may search documents attached to
the same stay. It does not define the output grain and it does not affect the
model's rationale. Later, `from_channel(group_by = "EVTID")` independently makes
one structured result record per stay and at most one model call when candidates
exist.

## 3. Put field meaning in an Ellmer schema

The model response is a native `ellmer::TypeObject`. Its object description
states the purpose of the record; each field description states how that field
must be interpreted.

```{r response-schema}
allergy_response <- ellmer::type_object(
  paste(
    "Extraction structuree du statut d'allergie medicamenteuse",
    "explicitement documente pour un sejour."
  ),
  statut_allergique = ellmer::type_enum(
    c(
      "allergie_documentee",
      "absence_allergie_documentee",
      "indetermine"
    ),
    paste(
      "Statut explicitement documente dans les extraits.",
      "Ne jamais deduire une absence d'allergie du silence."
    )
  ),
  substance_documentee = ellmer::type_string(
    paste(
      "Substance nommee comme allergene medicamenteux.",
      "Retourner 'non_documentee' si aucune substance n'est nommee."
    )
  )
)
```

The three instruction layers have separate jobs:

| Instruction layer | Responsibility |
|---|---|
| package system prompt | General rules for grounded structured extraction; override with `system_prompt =` only when the whole activation needs a different role |
| engine user prompt | Target keys plus numbered excerpts; `user_prompt =` adds an optional prefix for cross-field instructions |
| `TypeObject` and `type_*()` descriptions | Meaning, allowed values, and decision rules for this particular record and its fields |

There is no `prompt =` argument and no manually authored JSON format. The engine
calls `chat_structured()` with the TypeObject.

Do not add `rationale` or `evidence_ids` to the authored TypeObject. They are
engine-owned fields:

- omitting `rationale` in `use_channel()` adds one required rationale to the same
  structured response; a string overrides its description, while `FALSE` or
  `NULL` removes it;
- `evidence_ids` is injected at runtime as an enum containing only the snippet
  identifiers actually shown to the model.

The model therefore returns the authored fields, rationale, and citations in one
structured call. The engine resolves those citations into the public evidence
table and does not publish `evidence_ids` as a value column.

## 4. Declare the LLM activation and output record

```{r llm-variable}
medication_allergy_status <- variable_spec(
  name = "medication_allergy_status",
  concept = medication_allergy,
  channels = list(
    allergy_text = use_channel(
      channel = "allergy_text",
      search_within = "EVTID",
      method = "lucene_llm",
      model = "gemma3:4b",
      model_params = list(temperature = 0, seed = 42),
      response = allergy_response,
      max_candidates = 6L
    )
  ),
  output = from_channel(
    channel = "allergy_text",
    group_by = "EVTID"
  )
)

medication_allergy_status
```

`model` and `model_params` belong to this activation because they configure how
this particular LLM channel is processed. `gemma3:4b` is the model currently
accepted by the package's structured-output gate; it is an executable example,
not a general model recommendation.

Unlike deterministic `from_channel()`, an LLM output has no `value =`
expression. It publishes the complete structured frame: every authored field
plus the rationale added by the engine. Here `group_by = "EVTID"` constructs one
task, one model call when candidates exist, and one final result row per stay.

Printing or inspecting the specification resolves its authoring contract but
does not start Ollama or call the model.

## 5. Run the real model interactively

The following chunk is not evaluated when the vignette is built. To run it,
start Ollama locally and make sure `gemma3:4b` is available, then execute the code
in an interactive R session.

```{r real-llm-run, eval = FALSE}
allergy_run <- run_variable(
  medication_allergy_status,
  cohort = cohort,
  sources = list(documents = documents)
)
```

Because the model is declared in `use_channel()`, the engine constructs the
Ollama Chat and uses `chat_structured()` behind the scenes. `run_variable(chat =)`
can supply a real Ellmer Chat as a run-wide test or debugging override; it is not
needed for this normal path.

## 6. Read the result as four coordinated views

### Published record

```{r read-llm-values, eval = FALSE}
allergy_run$values |>
  select(
    PATID, EVTID,
    statut_allergique, substance_documentee, rationale,
    channel_coverage, needs_review, review_reason, citation_warning
  )
```

`values` always follows the declared output key. A task with no candidate or no
valid grounded response remains present with typed missing response fields and a
separate review state; it does not silently disappear. A no-match task normally
has `needs_review = FALSE` because selection completed without a model call.
Invalid or failed processing requires review. `citation_warning` distinguishes a
published grounded record that also contained unsupported citation IDs; only the
IDs that resolve to supplied snippets are materialized as evidence.

### Selection and model processing

```{r read-llm-status, eval = FALSE}
allergy_run$channel_status |>
  select(
    PATID, EVTID, channel, source,
    selection_status, processing_status
  )
```

The two status columns answer different questions:

| Situation | `selection_status` | `processing_status` | Interpretation |
|---|---|---|---|
| candidate and grounded response | `matched` | `completed` | retrieval found input and the model returned a publishable grounded record |
| no Lucene candidate | `no_match` | `not_called` | selection completed with no input, so no model call exists |
| no eligible document | `unavailable` | `not_called` | the retrieval boundary could not ascertain candidate presence |
| candidate but ungrounded response | `matched` | `invalid` | the call returned a structured response, but citation grounding did not permit publication |
| candidate but call or parsing failed | `matched` | `failed` | retrieval succeeded, but model transport or response processing failed |

These are mechanical execution states, not clinical labels. In particular,
`no_match` means *no excerpt matched the authored retrieval rule*, not *no allergy*.

### Grounded evidence

```{r read-llm-evidence, eval = FALSE}
allergy_run$evidence |>
  select(
    PATID, EVTID, source_EVTID,
    evidence_kind, evidence_ref, snippet_id, snippet_text
  )
```

For a successful LLM record, `evidence` contains only snippets cited by the
model, with `evidence_kind = "llm_citation"`. `EVTID` is the target output stay;
`source_EVTID` is the stay native to the cited document. They often have the
same value in event-scoped retrieval, but they have different roles.
`evidence_ref` is an opaque coordinate local to this run and source snapshot.

### Execution audit

```{r read-llm-audit, eval = FALSE}
allergy_run$audit$counts |>
  select(PATID, EVTID, channel, stage, unit, n)

allergy_run$audit$llm_calls |>
  select(
    PATID, EVTID, channel, model,
    call_status, response_status, task_validity,
    transport_attempts, latency_ms
  )

allergy_run$audit$execution_manifest
```

For text LLM channels, `counts` can expose searchable documents before and after
a window, Lucene matches, matches after activation filters, and snippets supplied
to the model. Stages appear only when the executor records a distinct count.

`llm_calls` has one row for each invocation actually attempted. A no-candidate
task has no call row. The call table separates transport outcome, response
processing, and final task validity; the execution manifest records the resolved
query, scope, model, schema, prompt configuration, and output grain.

## 7. Grounding is part of the publication contract

At runtime the schema's `evidence_ids` enum is built from the snippets visible in
that task's prompt. A response is publishable only when at least one returned ID
resolves to one of those supplied snippets. A response with no real citation is
marked invalid, its structured fields are not accepted into the study variable,
and the task remains visible for review.

This contract does not prove that the model's interpretation is clinically
correct. It provides a narrower guarantee: every published record is attached to
inspectable text that the model identified from the bounded evidence it saw.

## 8. One current boundary

An LLM response is a structured payload, not an implicit boolean hit. A
`lucene_llm` activation can be published with `from_channel()`, but it cannot
currently appear inside `combine_channels()`. The engine would need an explicit
authored rule mapping response fields to membership before that operation could
have an unambiguous meaning. Use `method = "lucene"` when Lucene-hit presence is
itself the intended signal.

## 9. Authoring checklist

Before running a new LLM variable, write down:

1. the query that defines candidate text and how you checked its hits;
2. whether documents are searched within `PATID` or `EVTID`;
3. the output `group_by`, which defines one target and at most one model call
   when candidates exist;
4. every response field, its allowed values, and its silence/uncertainty rule;
5. whether the default rationale is appropriate; and
6. which result, status, evidence, and audit columns you will inspect before
   accepting the variable.

Keep those decisions in the specification. The model supplies a structured
answer; the researcher still owns the operational definition and its validation.
